A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950

### 2022-02-02 paper.Rmd
updates

---
title: "A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950"
author: "Göran Broström and Tommy Bengtsson"
date: "`r Sys.time()`"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
  bookdown::word_document2:
    toc: no
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
citation_package: natbib
classoption: titlepage
bibliography: bio.bib
titlepage: yes
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{a4wide}
- \usepackage[figuresonly]{endfloat}
- \renewcommand{\efloatseparator}{\mbox{}}
abstract: "The effect of extreme temperatures on infant mortality in the Umeå region 1895-1950 is studied in a biometric analysis setting. More precisely, the effect of climate and weather, measured by temperature, on infant mortality is investigated. It turns out that climate is more important than weather, low average temperatures are more important than temporary dips in temperature. The investigated geographical region is situated too far north for bad effects of high temperatures to show."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=3,comment=NA)
library(knitr)
library(eha)
g <- function(x) (log(1 + x))^3 # x in days.
dg <- function(x) (3 * (log(1 + x))^2) / (1 + x)
ginv <- function(y) exp(y^(1/3)) - 1
```

@@

```{r mortstat, fig.cap = "Crude infant mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 4}
infdat <- readRDS("data/infdat3_year3.rds")
tid <- toTpch(Surv(enter, exit, event) ~ week, data = infdat, cuts = c(0, 1))
y <- tid$event / tid$exposure
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
     xlab = "Week", ylab = "IMR", cex = 0.7, lwd = 1.5)
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, las = 1, at = c(0, 0.03, 0.06, 0.09))
abline(h = c(0.05, 0.07, 0.09, 0.11), lty = 3)
abline(h = 0)
box()
```

@@

```{r neodata, fig.cap = "Crude neonatal mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 4}
neo <- age.window(infdat, c(0, 28 / 365))
tneo <- toTpch(Surv(enter, exit, event) ~ week, data = neo, cuts = c(0, 28 / 365))
y <- tneo$event / tneo$exposure
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
     xlab = "Week", ylab = "NMR", cex = 0.7, lwd = 1.5)
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, las = 1, at = c(0, 0.2, 0.4, 0.6))
abline(h = c(0.3, 0.4, 0.5, 0.6), lty = 3)
abline(h = 0)
box()
```

@@

```{r pneodata, fig.cap = "Crude postneonatal mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 4}
pneo <- age.window(infdat, c(28 / 365, 1))
tpneo <- toTpch(Surv(enter, exit, event) ~ week, data = pneo, cuts = c(28 / 365, 1))
y <- tpneo$event / tpneo$exposure
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
     xlab = "Week", ylab = "PNMR", cex = 0.7, lwd = 1.5)
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, at = c(0, 0.02, 0.04, 0.06), las = 1)
abline(h = c(0.03, 0.04, 0.05, 0.06, 0.07), lty = 3)
abline(h = 0)
box()
```

@@

```{r longtrend, fig.cap = "Crude IMR by year, Umeå-Skellefteå 1895--1950.", fig.height = 4}
trend <- toTpch(Surv(enter, exit, event) ~ year, data = infdat, cuts = c(0, 1))
y <- trend$event / trend$exposure
par(las = 1)
plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
     ylab = "IMR", lwd = 1.5, cex = 0.7)
abline(h = c(0.05, 0.1, 0.15), lty = 3)
abline(h = 0)
```

```{r ntrend, fig.cap = "Crude NMR by year, Umeå-Skellefteå 1895--1950.", fig.height = 4}
ndat <- age.window(infdat, c(0, 28 / 365))
tnda <- toTpch(Surv(enter, exit, event) ~ year, data = ndat, cuts = c(0, 28 / 365))
y <- tnda$event / tnda$exposure
par(las = 1)
plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
     ylab = "NMR", lwd = 1.5, cex = 0.7)
abline(h = 0)
abline(h = c(0.2, 0.4, 0.6), lty = 3)
```

@@

Weekly averages are calculated year by year, and deviations from 
the averages of the weekly averages are used as a time-varying *communal covariate*. See Figure 
\@ref(fig:monavgtemp) for the monthly distributions.

```{r monavgtemp, fig.cap="Weekly max and min temperature averages, 1895--1950.", fig.height = 4}
oldpar <- par(mfrow = c(1, 2))
## Umeå:
umaxavg <- umetemp$emaxtemp[1:52]
uminavg <- umetemp$emintemp[1:52]
plot(1:52, umaxavg, ylim = c(-20, 30), type = "b", col = "red", axes = FALSE,
     xlab = "Week", ylab = "Temperature (C)", cex = 0.6, lwd = 0.7)
text(17, 27, "Umeå")
axis(1, at = c(1, 10, 29, 43, 52))
axis(2, las = 1)
abline(h = 0)
abline(v = 29, lty = 3)
abline(h = c(-20, -10, 10, 20), lty = 3)
lines(1:52, uminavg, type = "b", col = "blue", cex = 0.6, lwd = 0.7)
text(8, 11, "max", col = "red")
text(20, -9, "min", col = "blue")
box()
## Skellefteå:
bmaxavg <- bjurtemp$emaxtemp[1:52]
bminavg <- bjurtemp$emintemp[1:52]
plot(1:52, bmaxavg, ylim = c(-20, 30), type = "b", col = "red", axes = FALSE,
     xlab = "Week", ylab = "Temperature (C)", cex = 0.6, lwd = 0.7)
text(17, 27, "Skellefteå")
axis(1, at = c(1, 10, 29, 43, 52))
axis(2, las = 1)
abline(h = 0)
abline(v = 29, lty = 3)
abline(h = c(-20, -10, 10, 20), lty = 3)
lines(1:52, bminavg, type = "b", col = "blue", cex = 0.6, lwd = 0.7)
text(8, 11, "max", col = "red")
text(20, -9, "min", col = "blue")
box()
##
par(oldpar)
```

Time trends of yearly average temperatures, see Figure

```{r longterm}
bmtemp <- rbind(umetemp, bjurtemp)
umt <- with(bmtemp, tapply(meantemp, year, mean))
plot(1894:1951, umt, type = "b", col = "blue", xlab = "Year",
     ylab = "Mean temperature", lwd = 1.5, cex = 0.7, ylim = c(0, 7))
abline(h = 0)
abline(h = 1:6, lty = 3)
```

## Mortality and temperature as a communal covariate

The two data sets, mortality and weather, are combined ino one by treating temperature data 
as a communal covariate and incorporate it as such in the mortality data set.

@@

The analyses are performed on the log-cube scale with proportional hazards modelling. Note that
the property of proportional hazards are preserved under a strictly monotone increasing time transform.
However, the estimates of baseline distribution characteristics will change, of course.

We start by performing separate analyses for neonatal and postneonatal mortality. Then we make comparisons.

@@

```{r postanalske, results='asis'}
source("R/ltx2.R")
infdat$parity <- cut(infdat$parity, c(-2, 1.5, 4.5, 19), labels = c("1", "2-4", "5+"))
infdat$period <- cut(infdat$birthdate, c(1895, 1914, 1935, 1951), dig.lab = 5)
levels(infdat$subreg) <- c("ume", "ske")
infdat$season <- cut(infdat$week, c(-1, 9.5, 22.5, 35.5, 48.5, 52.5), labels = c("winter", "spring", "summer", "fall", "winter"))
infdat$season <- as.factor(infdat$season)
pneo <- age.window(infdat, c(28 / 365, 1))
##
pneo$enter <- g(pneo$enter * 365)
pneo$exit <- g(pneo$exit * 365)
##
pneo$emintemp <- round(pneo$emintemp)
##
tpneo <- toTpch(Surv(enter, exit, event) ~ period + sex + socBranch + socStatus + illeg + parity + 
                    cold 
                + emintemp + season + subreg, 
                data = pneo, cuts = g(c(28, 365)))
##fit2 <- phreg(Surv(enter - g(28), exit - g(28), event) ~ sex + socBranch + socStatus + cold + emintemp, 
  ##           data = pneo, shape = 1)
##fit.pch <- pchreg(Surv(enter, exit, event) ~ sex + socBranch + socStatus + cold + emintemp, 
  ##           data = pneo, cuts = g(c(28, 365)))
tpneo$excessCold <- with(tpneo, -cold * (cold < 0))
fit.tpch <- tpchreg(oe(event, exposure) ~ excessCold + emintemp  +  period + sex + socBranch + socStatus  + 
                        parity + season, 
                    data = tpneo[tpneo$subreg == "ske", ], time = age)
dr <- drop1(fit.tpch, test = "Chisq")
ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, Skellefteå. Adjusted for social status and branch, illegitimacy, parity, time period, season, and sub region.", label = "tab:postanalske", keep = 2)
```

The result in Table \@ref(tab:postanalske) shows that *climate* (`emintemp`) is more important than 
*weather* (`cold`). Moreover, no signs of interaction between weather or climate and the rest of 
covariates (not shown).

```{r postanalume, results = 'asis'}
fit.tpch <- tpchreg(oe(event, exposure) ~ excessCold + emintemp + period + sex + socBranch + socStatus + 
                        parity + season, 
                    data = tpneo[tpneo$subreg == "ume", ], time = age)
dr <- drop1(fit.tpch, test = "Chisq")
ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, Umeå. Adjusted for social status and branch, illegitimacy, parity, time period, season, and subregion.", label = "tab:postanalume", keep = 2)
```

@@

```{r neoanal, results = 'asis'}
##library(survival)
neo <- age.window(infdat, c(0, 28 /365))
neo$exit <- g(neo$exit * 365)
neo$enter <- g(neo$enter * 365)
fit.ph <- coxreg(Surv(enter, exit, event) ~  cold + emintemp + socStatus + illeg + parity + period + socBranch + subreg,
                 data = neo[neo$sex == "girl", ])
dr <- drop1(fit.ph, test = "Chisq")
ltx2(fit.ph, dr = dr, caption = "Neonatal mortality, girls. Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", label = "tab:neonatal", keep = 2)
```

@@

```{r neoanalboys, results = 'asis'}
fit.ph <- coxreg(Surv(enter, exit, event) ~  cold + emintemp + socStatus + illeg + parity + period + socBranch + subreg,
                 data = neo[neo$sex == "boy", ])
dr <- drop1(fit.ph, test = "Chisq")
ltx2(fit.ph, dr = dr, caption = "Neonatal mortality, boys. Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", label = "tab:neonatalboys", keep = 2)
```

@@

```{r postanalw, results='asis'}
##infdat$parity <- cut(infdat$parity, c(-2, 1.5, 4.5, 19), labels = c("1", "2-4", "5+"))
##levels(infdat$subreg) <- c("ume", "ske")
##infdat$season <- cut(infdat$week, c(-1, 9.5, 22.5, 35.5, 48.5, 52.5), labels = c("winter", "spring", "summer", "fall", "winter"))
##infdat$season <- as.factor(infdat$season)
pneo <- age.window(infdat[infdat$season %in% c("winter", "spring"), ], c(28 / 365, 1))
##
pneo$enter <- g(pneo$enter * 365)
pneo$exit <- g(pneo$exit * 365)
##
pneo$emintemp <- round(pneo$emintemp)
##
tpneo <- toTpch(Surv(enter, exit, event) ~ sex + socBranch + socStatus + illeg + parity + 
                    cold + emintemp + period + subreg, 
                data = pneo, cuts = g(c(28, 365)))
##fit2 <- phreg(Surv(enter - g(28), exit - g(28), event) ~ sex + socBranch + socStatus + cold + emintemp, 
  ##           data = pneo, shape = 1)
##fit.pch <- pchreg(Surv(enter, exit, event) ~ sex + socBranch + socStatus + cold + emintemp, 
  ##           data = pneo, cuts = g(c(28, 365)))
fit.tpch <- tpchreg(oe(event, exposure) ~ cold + emintemp + sex + socBranch + socStatus + illeg + parity + period + 
                        subreg, 
                    data = tpneo, time = age)
dr <- drop1(fit.tpch, test = "Chisq")
ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring. Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", 
     label = "tab:postanalw", keep = 2)
```

@@

```{r neoanalw, results = 'asis'}
neo <- age.window(infdat[infdat$season %in% c("winter", "spring"), ], c(0, 28 /365))
neo$exit <- g(neo$exit * 365)
neo$enter <- g(neo$enter * 365)
fit.ph <- coxreg(Surv(enter, exit, event) ~  cold + emintemp + socStatus + illeg + parity + period + socBranch + subreg,
                 data = neo[neo$sex == "girl", ])
dr <- drop1(fit.ph, test = "Chisq")
ltx2(fit.ph, dr = dr, caption = "Neonatal mortality winter and spring, girls. Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", label = "tab:neonatalw", keep = 2)
```

@@

```{r neoanalboysw, results = 'asis'}
fit.ph <- coxreg(Surv(enter, exit, event) ~  cold + emintemp + socStatus + illeg + parity + period + socBranch + subreg,
                 data = neo[neo$sex == "boy", ])
dr <- drop1(fit.ph, test = "Chisq")
ltx2(fit.ph, dr = dr, caption = "Neonatal mortality winter and spring, boys. Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", 
     label = "tab:neonatalboysw", keep = 2)
```

>>>>

---
title: "A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950"
author: "Göran Broström and Tommy Bengtsson"
date: "`r Sys.time()`"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
  bookdown::word_document2:
    toc: no
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
citation_package: natbib
classoption: titlepage
bibliography: bio.bib
titlepage: yes
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{a4wide}
#- \usepackage[figuresonly]{endfloat}
#- \renewcommand{\efloatseparator}{\mbox{}}
abstract: "The effect of extreme temperatures on infant mortality in the Umeå region 1895-1950 is studied in a biometric analysis setting. More precisely, the effect of climate and weather, measured by temperature, on infant mortality is investigated. It turns out that climate is more important than weather, low average temperatures are more important than temporary dips in temperature. The investigated geographical region is situated too far north for bad effects of high temperatures to show."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=3, comment=NA, global.par=TRUE)
library(knitr)
library(eha)
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
g <- function(x) (log(1 + x))^3 # x in days.
dg <- function(x) (3 * (log(1 + x))^2) / (1 + x)
ginv <- function(y) exp(y^(1/3)) - 1
```

@@

```{r mortstat, fig.cap = "Crude infant mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3}
infdat <- readRDS("data/infdat3_year3.rds")
tid <- toTpch(Surv(enter, exit, event) ~ week, data = infdat, cuts = c(0, 1))
y <- tid$event / tid$exposure
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
     xlab = "Week", ylab = "IMR", cex = 0.7, lwd = 1.5)
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, las = 1, at = c(0, 0.03, 0.06, 0.09))
abline(h = c(0.05, 0.07, 0.09, 0.11), lty = 3)
abline(h = 0)
box()
```

@@

```{r neodata, fig.cap = "Crude neonatal mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3}
neo <- age.window(infdat, c(0, 28 / 365))
tneo <- toTpch(Surv(enter, exit, event) ~ week, data = neo, cuts = c(0, 28 / 365))
y <- tneo$event / tneo$exposure
par(cex.axis = 0.7)
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
     xlab = "Week", ylab = "NMR", cex = 0.7, lwd = 1.5)
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, las = 1, at = c(0, 0.2, 0.4, 0.6))
abline(h = c(0.3, 0.4, 0.5, 0.6), lty = 3)
abline(h = 0)
box()
```

@@

```{r pneodata, fig.cap = "Crude postneonatal mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3}
pneo <- age.window(infdat, c(28 / 365, 1))
tpneo <- toTpch(Surv(enter, exit, event) ~ week, data = pneo, cuts = c(28 / 365, 1))
y <- tpneo$event / tpneo$exposure
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
     xlab = "Week", ylab = "PNMR", cex = 0.7, lwd = 1.5)
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, at = c(0, 0.02, 0.04, 0.06), las = 1)
abline(h = c(0.03, 0.04, 0.05, 0.06, 0.07), lty = 3)
abline(h = 0)
box()
```

@@

```{r longtrend, fig.cap = "Crude IMR by year, Umeå-Skellefteå 1895--1950.", fig.height = 3}
trend <- toTpch(Surv(enter, exit, event) ~ year, data = infdat, cuts = c(0, 1))
y <- trend$event / trend$exposure
par(las = 1)
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
     ylab = "IMR", lwd = 1.5, cex = 0.7)
abline(h = c(0.05, 0.1, 0.15), lty = 3)
abline(h = 0)
```

```{r ntrend, fig.cap = "Crude NMR by year, Umeå-Skellefteå 1895--1950.", fig.height = 4}
ndat <- age.window(infdat, c(0, 28 / 365))
tnda <- toTpch(Surv(enter, exit, event) ~ year, data = ndat, cuts = c(0, 28 / 365))
y <- tnda$event / tnda$exposure
par(las = 1)
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
     ylab = "NMR", lwd = 1.5, cex = 0.7)
abline(h = 0)
abline(h = c(0.2, 0.4, 0.6), lty = 3)
```

@@

Weekly averages (`mintemp`, `maxtemp`, `meantemp`)  are calculated by week and year, and deviations from 
the averages (`emintemp`, `emaxtemp`, `emeantemp`) of the weekly averages are used as time-varying *communal covariates*. 
As an example, see Figure \@ref(fig:firstweek), where the variation around the average
minimum temperature (`emintemp`) week 1 is shown.

```{r firstweek, fig.cap = "Minimum temperature the first week of each year.", fig.height = 3}
w1 <- umetemp[umetemp$week == 1, ]
emintemp <- w1$emintemp[1]
mintemp <- w1$mintemp
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(w1$year, mintemp, type = "b", lty = 2, col = "blue", xlab = "Year", ylab = "Temperature", 
     axes = FALSE, ylim = c(-35, 0), cex = 0.7)
axis(1, at = c(1894, 1903, 1913, 1923, 1932, 1942, 1951))
whh <- c(-30, -25, emintemp, -10, 0) 
axis(2, at = whh, labels = round(whh))
abline(h = emintemp, lwd = 1.5, col = "blue")
abline (h = 0)
abline(v = 1923, lty = 3)
box()
```

Curiously, our randomly selected year 1923 turns out contain the warmest first week of all years, see 
Figure \@ref(fig:firstweek).

Figure 
\@ref(fig:monavgtemp) shows the average monthly distribution over all years. The subregional patterns 
and levels are very similar.

```{r monavgtemp, fig.cap="Weekly max and min temperature averages, 1895--1950.", fig.height = 3}
oldpar <- par(mfrow = c(1, 2))
## Umeå:
umaxavg <- umetemp$emaxtemp[1:52]
uminavg <- umetemp$emintemp[1:52]
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1:52, umaxavg, ylim = c(-20, 30), type = "b", col = "red", axes = FALSE,
     xlab = "Week", ylab = "Temperature (C)", cex = 0.6, lwd = 0.7)
text(17, 27, "Umeå", cex = 0.7)
axis(1, at = c(1, 10, 29, 43, 52))
axis(2, las = 1)
abline(h = 0)
abline(v = 29, lty = 3)
abline(h = c(-20, -10, 10, 20), lty = 3)
lines(1:52, uminavg, type = "b", col = "blue", cex = 0.6, lwd = 0.7)
text(8, 11, "max", col = "red", cex = 0.7)
text(20, -9, "min", col = "blue", cex = 0.7)
box()
## Skellefteå:
bmaxavg <- bjurtemp$emaxtemp[1:52]
bminavg <- bjurtemp$emintemp[1:52]
plot(1:52, bmaxavg, ylim = c(-20, 30), type = "b", col = "red", axes = FALSE,
     xlab = "Week", ylab = "Temperature (C)", cex = 0.6, lwd = 0.7)
text(17, 27, "Skellefteå", cex = 0.7)
axis(1, at = c(1, 10, 29, 43, 52))
axis(2, las = 1)
abline(h = 0)
abline(v = 29, lty = 3)
abline(h = c(-20, -10, 10, 20), lty = 3)
lines(1:52, bminavg, type = "b", col = "blue", cex = 0.6, lwd = 0.7)
text(8, 11, "max", col = "red", cex = 0.7)
text(20, -9, "min", col = "blue", cex = 0.7)
box()
##
par(oldpar)
```

Time trends of yearly average temperatures, see Figure \@ref(fig:longterm).

```{r longterm, fig.cap = "Yearly average temperatures, Umeå and Skellefteå."}
bmtemp <- rbind(umetemp, bjurtemp)
umt <- with(bmtemp, tapply(meantemp, year, mean))
par(las = 1)
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1894:1951, umt, type = "b", col = "blue", xlab = "Year",
     ylab = "Mean temperature", lwd = 1.5, cex = 0.7, ylim = c(0, 7))
abline(h = 0)
abline(h = 1:6, lty = 3)
```

## Temperature as communal covariates

The two data sets, mortality and weather, are combined into one by treating temperature data 
as a communal covariate and incorporate it as such in the mortality data set. The 
function *make.communal* in the **R** [@cran] package *eha* [@eha; @ehar2] is used 
for that purpose.

```{r readdata3}
levels(infdat$subreg) <- c("ume", "ske")
##infdat$excessTemp <- cut(infdat$aver, c(-16, -2, 2, 9), include.lowest = TRUE) # NOTE!!!
##infdat$excessTemp <- relevel(infdat$excessTemp, ref = "(-2,2]")
infdat$excessTemp <- infdat$cold # CHECK THIS!!
tbl(head(infdat))
```

@@

The analyses are performed on the *log-cube* scale, following the hints of  
Bourgeois-Pichat [@bp51a; @bp51b], with proportional hazards modelling. Note that
the property of proportional hazards are preserved under a strictly monotone increasing time transform.
However, the estimates of baseline distribution characteristics will change, of course.

We start by performing separate analyses for neonatal and postneonatal mortality.

@@

```{r postanal, results='asis'}
source("R/ltx2.R")
infdat$parity <- cut(infdat$parity, c(-2, 1.5, 4.5, 19), labels = c("1", "2-4", "5+"))
infdat$period <- cut(infdat$birthdate, c(1895, 1914, 1935, 1951), dig.lab = 5)
levels(infdat$subreg) <- c("ume", "ske")
infdat$season <- cut(infdat$week, c(-1, 9.5, 22.5, 35.5, 48.5, 52.5), labels = c("winter", "spring", "summer", "fall", "winter"))
infdat$season <- as.factor(infdat$season)
pneo <- age.window(infdat, c(28 / 365, 1))
##
pneo$enter <- g(pneo$enter * 365)
pneo$exit <- g(pneo$exit * 365)
##
pneo$emintemp <- round(pneo$emintemp)
##
tpneo <- toTpch(Surv(enter, exit, event) ~ period + sex + socBranch + socStatus + illeg + parity + 
                    excessTemp
                + emintemp + season + subreg, 
                data = pneo, cuts = g(c(28, 365)))
##fit2 <- phreg(Surv(enter - g(28), exit - g(28), event) ~ sex + socBranch + socStatus + cold + emintemp, 
  ##           data = pneo, shape = 1)
##fit.pch <- pchreg(Surv(enter, exit, event) ~ sex + socBranch + socStatus + cold + emintemp, 
  ##           data = pneo, cuts = g(c(28, 365)))

fit.tpch <- tpchreg(oe(event, exposure) ~ excessTemp + emintemp +  period + sex + socBranch + socStatus  + 
                        illeg + parity, 
                    data = tpneo, time = age)
dr <- drop1(fit.tpch, test = "Chisq")
ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality. Adjusted for social status and branch, illegitimacy, parity, time period, season, and subregion.", label = "tab:postanalske", keep = 2)
```

The result in Table \@ref(tab:postanal) shows that *climate* (`emintemp`) is more important than 
*weather* (`excessTemp`). Moreover, no signs of interaction between weather or climate and the rest of 
covariates (not shown).

@@

```{r neoanal, results = 'asis'}
##library(survival)
neo <- age.window(infdat, c(0, 28 /365))
neo$exit <- g(neo$exit * 365)
neo$enter <- g(neo$enter * 365)
fit.ph <- coxreg(Surv(enter, exit, event) ~  (excessTemp + emintemp) * sex + socStatus + illeg + parity + period + socBranch + subreg,
                 data = neo)
dr <- drop1(fit.ph, test = "Chisq")
ltx2(fit.ph, dr = dr, caption = "Neonatal mortality. Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", label = "tab:neonatal", keep = 2)
```

@@

```{r neoanalboys, results = 'asis'}
fit.ph <- coxreg(Surv(enter, exit, event) ~  excessTemp + emintemp + socStatus + illeg + parity + period + socBranch + subreg,
                 data = neo[neo$sex == "boy" & neo$season %in% c("spring", "summer"), ])
dr <- drop1(fit.ph, test = "Chisq")
ltx2(fit.ph, dr = dr, caption = "Neonatal mortality, boys. Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", label = "tab:neonatalboys", keep = 2)
```

@@

```{r postanalw, results='asis'}
##infdat$parity <- cut(infdat$parity, c(-2, 1.5, 4.5, 19), labels = c("1", "2-4", "5+"))
##levels(infdat$subreg) <- c("ume", "ske")
##infdat$season <- cut(infdat$week, c(-1, 9.5, 22.5, 35.5, 48.5, 52.5), labels = c("winter", "spring", "summer", "fall", "winter"))
##infdat$season <- as.factor(infdat$season)
pneo <- age.window(infdat[infdat$season %in% c("winter", "spring"), ], c(28 / 365, 1))
##
pneo$enter <- g(pneo$enter * 365)
pneo$exit <- g(pneo$exit * 365)
##
pneo$emintemp <- round(pneo$emintemp)
##
tpneo <- toTpch(Surv(enter, exit, event) ~ excessTemp + emintemp + sex + socBranch + socStatus + illeg + parity + 
                    period + subreg, 
                data = pneo, cuts = g(c(28, 365)))
##fit2 <- phreg(Surv(enter - g(28), exit - g(28), event) ~ sex + socBranch + socStatus + cold + emintemp, 
  ##           data = pneo, shape = 1)
##fit.pch <- pchreg(Surv(enter, exit, event) ~ sex + socBranch + socStatus + cold + emintemp, 
  ##           data = pneo, cuts = g(c(28, 365)))
fit.tpch <- tpchreg(oe(event, exposure) ~ excessTemp + emintemp + sex + socBranch + socStatus + illeg + parity + period + 
                        subreg, 
                    data = tpneo, time = age)
dr <- drop1(fit.tpch, test = "Chisq")
ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring. Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", 
     label = "tab:postanalw", keep = 2)
```

@@

```{r neoanalw, results = 'asis'}
neo <- age.window(infdat[infdat$season %in% c("winter", "spring"), ], c(0, 28 /365))
neo$exit <- g(neo$exit * 365)
neo$enter <- g(neo$enter * 365)
fit.ph <- coxreg(Surv(enter, exit, event) ~  excessTemp + emintemp + socStatus + illeg + parity + period + socBranch + subreg,
                 data = neo[neo$sex == "girl" & neo$emeantemp < -0, ])
dr <- drop1(fit.ph, test = "Chisq")
ltx2(fit.ph, dr = dr, caption = "Neonatal mortality winter and spring, girls. Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", label = "tab:neonatalw", keep = 2)
```

@@

```{r neoanalboysw, results = 'asis'}
fit.ph <- coxreg(Surv(enter, exit, event) ~  excessTemp + emintemp + socStatus + illeg + parity + period + socBranch + subreg,
                 data = neo[neo$sex == "boy" & neo$season %in% c("winter", "spring"), ])
dr <- drop1(fit.ph, test = "Chisq")
ltx2(fit.ph, dr = dr, caption = "Neonatal mortality winter and spring, boys. Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", 
     label = "tab:neonatalboysw", keep = 2)
```
